---
title: "DATA 622 Project"
author: "Catherine Dube, Guillermo Schneider, Tai Chou-Kudu"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  out.width = "100%", 
  fig.align = "center"
)

if(!require(pacman)) install.packages("pacman")
pacman::p_load(
   tidyverse, 
   fpp3, 
   rstudioapi,
   tidycensus,
   r5r, 
   sf, #spacial data handling
   tigris, #NYC geography
   mapview,
   osmextract,
   arrow, # read parquet data, designed for random access?
   dplyr,
   knitr,
   janitor, # clean up the names of columns that have spaces (subway entrances data)
   tidygeocoder,
   broom
)
projPath <- dirname(file.path(getSourceEditorContext()$path))

```


# Modeling the Impact of the IBX on 60minute+ commute times 

## Foundation 

### Census Data 

Looking at census data from **2022**. 

Picked 2022 because:

* This is a good Post-Pandemic option; NYC was "fully reopened" 
* The 5-Year ACS helps avoid the 2020 collection gap (where due to COVID, the Census Bureau couldn't send people door to door) and the 2021 remote-forward living arrangements, where many people left Manhattan during the high period of remote work.
* The 5-year estimate is a rolling average (2018–2022), and some COVID-impacted data is included either way. However, multiple years of more robust data are included in this time window.

Another option would be to switch to the most recent census data available, but it's worth noting that that won't avoid the COVID data being included in the 5 -year estimate.


#### Current Vars from census

```{r acquire-census-data} 

#CENSUS_API_KEY is in environment 

nyc_counties <- c("New York", "Kings", "Queens", "Bronx", "Richmond")

 
# Entire 2022 5-Year ACS variable dictionary
acs_vars_2022 <- load_variables(2022, "acs5", cache = TRUE)

# MISC.
# Exploring different potential features
feature_exploration <- acs_vars_2022 %>%
  filter(
    str_detect(str_to_lower(label), "public transportation|bicycle|walked|worked from home") |
    str_detect(str_to_lower(concept), "means of transportation to work|travel time")
  )

# Comment Me out before final knit
#View(feature_exploration)
```

```{r more-census-data, results='hide', message=FALSE, cache=TRUE}
# hiding the results of this chunk becuase it jsut is unreadable in the knitted document

# Census variables
vars <- c(
  # Base Demographics
  no_car      = "B08201_002", 
  med_income  = "B19013_001", 
  med_age     = "B01002_001",
  total_pop   = "B01003_001",
  
  # Transportation Base & Features
  tot_workers = "B08301_001",
  transit     = "B08301_010",
  walk        = "B08301_019",
  wfh         = "B08301_021",
  
  # Travel Time Features
  # Potential target variable
  commute_60  = "B08303_012",
  commute_90  = "B08303_013",
  
  # Education Base & Features
  tot_pop_25  = "B15003_001",
  bachelors   = "B15003_022",
  
  # Occupation Base & The 5 Major Buckets
  tot_employ  = "C24010_001",
  mngmt_job   = "C24010_002",
  service_job = "C24010_014",
  sales_job   = "C24010_021",
  const_job   = "C24010_030",
  trans_job   = "C24010_034",

  # Continuous Age Spectrum (Table B01001 - Detailed Age)
  # 18 to 24
  m_18_19="B01001_007", m_20="B01001_008", m_21="B01001_009", m_22_24="B01001_010",
  f_18_19="B01001_031", f_20="B01001_032", f_21="B01001_033", f_22_24="B01001_034",
  # 25 to 34
  m_25_29="B01001_011", m_30_34="B01001_012", f_25_29="B01001_035", f_30_34="B01001_036",
  # 35 to 44
  m_35_39="B01001_013", m_40_44="B01001_014", f_35_39="B01001_037", f_40_44="B01001_038",
  # 45 to 64
  m_45_49="B01001_015", m_50_54="B01001_016", m_55_59="B01001_017", m_60_61="B01001_018", m_62_64="B01001_019",
  f_45_49="B01001_039", f_50_54="B01001_040", f_55_59="B01001_041", f_60_61="B01001_042", f_62_64="B01001_043",
  # 65 Plus
  m_65_66="B01001_020", m_67_69="B01001_021", m_70_74="B01001_022", m_75_79="B01001_023", m_80_84="B01001_024", m_85_up="B01001_025",
  f_65_66="B01001_044", f_67_69="B01001_045", f_70_74="B01001_046", f_75_79="B01001_047", f_80_84="B01001_048", f_85_up="B01001_049",

  # Commercial vs. Residential Proxy (Housing)
  tot_housing = "B25002_001",
  vacant_unit = "B25002_003",
  tot_occup   = "B25003_001",
  renter_unit = "B25003_003",

  # Economic Barrier Spectrum (Poverty)
  tot_pov_base = "B17001_001",
  pov_below    = "B17001_002"
) 

# Getting from census data
nyc_tracts <- get_acs(
  geography = "tract",
  variables = vars,
  state = "NY",
  county = nyc_counties,
  geometry = TRUE, 
  year = 2022,
  output = "wide"
)

```
#### Feature Engineering Census Data

This chunk does an explicit selection of columns from the census data to start with. (There are just so many features in that data that clearly selecting and renaming when applicable seemed like a cleaner route.)

After that this chunk is doing some feature engineering on the census data. Mostly turning raw count data into percent of population data.

* `pct_commute_60p` is currently being used as the target variable. It's the count of people with commutes over 60 min + the count of people with commutes over 90 minutes. Divided by total number of workers. x 100 to make is a percentage.
* `pct_no_car` is percent of people who do not have a car 
* `pct_transit` is percent of workers using public transit (mta, buses **not** uber/taxi) as primary commute methos 
* `pct_walk` is percent of workers using walking as primary commute method
* `pct_wfh`is percent of workers working from home 
* `pct_bachelors` is the percent of 25+ year olds with Bachelor's Degrees or higher
* Occupation buckets
 * `pct_mngmt` is percent of workers in management, business, science, and arts
 * `pct_service` is percent of workers in 
 * `pct_sales` is percent of workers in sales and office administration
 * `pct_const` is percent of workers in natural resources, construction, and maintenance
 * `pct_trans` is percent of workers in production, transportation, and material moving
* `area_sq_miles` is the area in square miles
* `pop_density`is the numebr of people per square mile
* Age GRoups
 * I didn't want to bucket the ages because I know having continuous data is better, but unfortuantely the census already had some underlying bucketing
 * `pct_age_18_24` is the percent of the population in that age range
 * `pct_age_25_34` is the percent of the population in that age range
 * `pct_age_35_44` is the percent of the population in that age range
 * `pct_age_45_64` is the percent of the population in that age range
 * `pct_age_65_plus` is the percent of the population in that age range
* `pct_vacant` is the percent of housing that is vacant
* `pct_renter` is the percent of occupued units that are being rented
* `pct_poverty` is the percent of people living below the poverty line out of the total population for whom poverty status is determined
    
Not all of the engineered features are used in the later model.

```{r feature-engineering-census-data}
nyc_tracts <- nyc_tracts %>%
  # Explicitly rename every variable 
  select(
    GEOID, NAME, geometry,
    
    # Base Demographics
    no_car      = no_carE, 
    med_income  = med_incomeE, 
    med_age     = med_ageE, 
    total_pop   = total_popE, 
    tot_workers = tot_workersE, 
    tot_pop_25  = tot_pop_25E, 
    tot_employ  = tot_employE,
    
    # Transportation & Travel Time
    transit     = transitE, 
    walk        = walkE, 
    wfh         = wfhE, 
    commute_60  = commute_60E, 
    commute_90  = commute_90E, 
    
    # Education & Occupation
    bachelors   = bachelorsE, 
    mngmt_job   = mngmt_jobE, 
    service_job = service_jobE, 
    sales_job   = sales_jobE, 
    const_job   = const_jobE, 
    trans_job   = trans_jobE,
    
    # Age - Male
    m_18_19 = m_18_19E, m_20 = m_20E, m_21 = m_21E, m_22_24 = m_22_24E,
    m_25_29 = m_25_29E, m_30_34 = m_30_34E,
    m_35_39 = m_35_39E, m_40_44 = m_40_44E,
    m_45_49 = m_45_49E, m_50_54 = m_50_54E, m_55_59 = m_55_59E, m_60_61 = m_60_61E, m_62_64 = m_62_64E,
    m_65_66 = m_65_66E, m_67_69 = m_67_69E, m_70_74 = m_70_74E, m_75_79 = m_75_79E, m_80_84 = m_80_84E, m_85_up = m_85_upE,
    
    # Age - Female
    f_18_19 = f_18_19E, f_20 = f_20E, f_21 = f_21E, f_22_24 = f_22_24E,
    f_25_29 = f_25_29E, f_30_34 = f_30_34E,
    f_35_39 = f_35_39E, f_40_44 = f_40_44E,
    f_45_49 = f_45_49E, f_50_54 = f_50_54E, f_55_59 = f_55_59E, f_60_61 = f_60_61E, f_62_64 = f_62_64E,
    f_65_66 = f_65_66E, f_67_69 = f_67_69E, f_70_74 = f_70_74E, f_75_79 = f_75_79E, f_80_84 = f_80_84E, f_85_up = f_85_upE,

    # Housing & Poverty
    tot_housing  = tot_housingE, 
    vacant_unit  = vacant_unitE, 
    tot_occup    = tot_occupE, 
    renter_unit  = renter_unitE, 
    tot_pov_base = tot_pov_baseE, 
    pov_below    = pov_belowE
  ) %>%
  
  # Feature Engineering: Convert raw counts to percentages
  mutate(
    pct_no_car      = (no_car / total_pop) * 100,
    pct_transit     = (transit / tot_workers) * 100,
    pct_walk        = (walk / tot_workers) * 100,
    pct_wfh         = (wfh / tot_workers) * 100,
    pct_commute_60p = ((commute_60 + commute_90) / tot_workers) * 100,
    pct_bachelors   = (bachelors / tot_pop_25) * 100,
    pct_mngmt       = (mngmt_job / tot_employ) * 100,
    pct_service     = (service_job / tot_employ) * 100,
    pct_sales       = (sales_job / tot_employ) * 100,
    pct_const       = (const_job / tot_employ) * 100,
    pct_trans       = (trans_job / tot_employ) * 100,
    area_sq_miles   = as.numeric(st_area(geometry)) / 2589988, 
    pop_density     = total_pop / area_sq_miles,
    
    # Continuous Age Spectrum
    pct_age_18_24   = ((m_18_19 + m_20 + m_21 + m_22_24 + f_18_19 + f_20 + f_21 + f_22_24) / total_pop) * 100,
    pct_age_25_34   = ((m_25_29 + m_30_34 + f_25_29 + f_30_34) / total_pop) * 100,
    pct_age_35_44   = ((m_35_39 + m_40_44 + f_35_39 + f_40_44) / total_pop) * 100,
    pct_age_45_64   = ((m_45_49 + m_50_54 + m_55_59 + m_60_61 + m_62_64 + f_45_49 + f_50_54 + f_55_59 + f_60_61 + f_62_64) / total_pop) * 100,
    pct_age_65_plus = ((m_65_66 + m_67_69 + m_70_74 + m_75_79 + m_80_84 + m_85_up + f_65_66 + f_67_69 + f_70_74 + f_75_79 + f_80_84 + f_85_up) / total_pop) * 100,
    
    # Commercial/Residential Proxy
    pct_vacant      = (vacant_unit / tot_housing) * 100,
    pct_renter      = (renter_unit / tot_occup) * 100,
    
    # Economic Barrier
    pct_poverty     = (pov_below / tot_pov_base) * 100
  ) %>%
  
  # Explicitly drop the raw counts so they don't leak into the model
  select(
    -no_car, -transit, -walk, -wfh, -commute_60, -commute_90, -bachelors, 
    -mngmt_job, -service_job, -sales_job, -const_job, -trans_job, 
    -tot_workers, -tot_pop_25, -tot_employ,
    -m_18_19, -m_20, -m_21, -m_22_24, -m_25_29, -m_30_34, -m_35_39, -m_40_44, 
    -m_45_49, -m_50_54, -m_55_59, -m_60_61, -m_62_64, -m_65_66, -m_67_69, -m_70_74, -m_75_79, -m_80_84, -m_85_up,
    -f_18_19, -f_20, -f_21, -f_22_24, -f_25_29, -f_30_34, -f_35_39, -f_40_44, 
    -f_45_49, -f_50_54, -f_55_59, -f_60_61, -f_62_64, -f_65_66, -f_67_69, -f_70_74, -f_75_79, -f_80_84, -f_85_up,
    -tot_housing, -vacant_unit, -tot_occup, -renter_unit, -tot_pov_base, -pov_below
  )
```

#### Checking for missing census data

```{r census-missing-data-exploration}
## Checking for total NAs and NaNs in each column
colSums(is.na(nyc_tracts %>% st_drop_geometry())) %>%
  kable(caption = "# of NAs and NaNs in each column")

## Percentage of missing data 
# (is it a low enough percent to be negligible) 
missing_summary <- nyc_tracts %>%
  st_drop_geometry() %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(cols = everything(), names_to = "Feature", values_to = "Missing_Count") %>%
  mutate(
    Total_Rows = nrow(nyc_tracts),
    Pct_Missing = (Missing_Count / Total_Rows) * 100,
    
    # Flag if the missing data is below 5% 
    Is_Negligible_Under_5Pct = Pct_Missing < 5 
  ) %>%
  # Filter to only show features that actually have missing data
  filter(Missing_Count > 0) %>% 
  arrange(desc(Pct_Missing))

# Output Percentage of Missing Data
kable(missing_summary, digits = 2, caption = "Missing Data Percentage by Feature")

## Map for missing income data
missing_map_data <- nyc_tracts %>%
  mutate(missing_income_flag = ifelse(is.na(med_income), "Missing Data", "Valid Data"))

ggplot(missing_map_data) +
  geom_sf(aes(fill = missing_income_flag), color = "white", size = 0.1) +
  scale_fill_manual(values = c("Valid Data" = "#e0e0e0", "Missing Data" = "#d73027")) +
  theme_void() +
  labs(title = "Map of Missing Census Income Data",
       fill = "Status") 
```

#### Census NA Handling

This chunk is currently dropping census tracts that have very few people or no workers.

It's also **imputing** where income or age was missing using the median value. This was just a simple implementation, but we can definitely consider more robust imputing methos, and also properly investingating MAR, MNAR etc. (Though given the nature of census data, MNAR is a good assumption.)

```{r census-na-handling}
## Drop tracts with less than 50 people or 0 workers
# There were a ton of tracts with a small number of or 0 people, screwing with denominators and also making for super volatile percentages
nyc_tracts_clean <- nyc_tracts %>%
  filter(total_pop >= 50, !is.na(pct_transit))

## Imputing
# There were also many tracts where income or age was missing, but it wasn't due to a denominator issue
# Just doing simple median imputation
# More robust method would be better
# Checking for MNAR and stuff would be better
# The current median imputation: 
nyc_tracts_clean <- nyc_tracts_clean %>%
  # Extract the borough from the NAME string (e.g., "Queens County")
  mutate(borough = str_extract(NAME, "(?<=; ).*(?= County)")) %>%
  
  # Group by Borough so the imputation is localized
  group_by(borough) %>%
  mutate(
    med_income = ifelse(is.na(med_income), 
                        median(med_income, na.rm = TRUE), 
                        med_income),
    med_age = ifelse(is.na(med_age), 
                     median(med_age, na.rm = TRUE), 
                     med_age)
  ) %>%
  ungroup() %>%
  
  # Drop the temporary borough column 
  select(-borough)
 
print(paste("Original tracts:", nrow(nyc_tracts)))
print(paste("Cleaned tracts of NAs:", nrow(nyc_tracts_clean)))
```


```{r save-census-data}
# Save 
census_save_path <- file.path(projPath, "data", "nyc_census_tracts.rds") 
saveRDS(nyc_tracts_clean, census_save_path)

print(paste("Census features and geometry saved to:", census_save_path))

# Peak
head(nyc_tracts_clean)
```

### Taxi Zones

The taxi zone data is luckily pretty straightforward. It's just being imported here.

```{r taxi-zones}  
#Zip file from below url is downloaded and extracted into repo
#taxi_zones_url <- "https://d37ci6vzurychx.cloudfront.net/misc/taxi_zones.zip" 

# Read the shapefile
taxi_zones_sf <- st_read(file.path(projPath, 'taxi_zones', 'taxi_zones', "taxi_zones.shp")) %>%
  st_transform(4326) # Transform to standard Lat/Long (WGS84)

# Lookup table with names and boroughs is downloaded into repo
#zone_lookup_url <- "https://d37ci6vzurychx.cloudfront.net/misc/taxi_zone_lookup.csv"
zone_lookup <- read_csv(file.path(projPath, 'taxi_zone_lookup.csv'))

# Join 
taxi_zones <- taxi_zones_sf %>%
  left_join(zone_lookup, by = c("LocationID" = "LocationID"))

# Peak: Does this map look right 
plot(st_geometry(taxi_zones))
```

### TLC Data

TLC Data is from the [NYC Taxi and Limousine Commission](https://www.nyc.gov/site/tlc/about/tlc-trip-record-data.page).

High volume for hire vehicles = Uber/Lyft.

* Originally was pulling **January 2026** data for High Volume for-hire vehicles.
 * https://d37ci6vzurychx.cloudfront.net/trip-data/fhvhv_tripdata_2026-01.parquet 
* Currently pulling **October 2023** data for yellow taxis, green taxis, ~for-hire vehicles (FHV (local livery/black car)),~ and high volume for-hire vehicles (FHVHV (Uber/Lyft) Uber/Lyft). This made a little more sense with the 2022 census data, and with the seasonality of commuting and to avoid holiday season. 
 * https://d37ci6vzurychx.cloudfront.net/trip-data/yellow_tripdata_2023-10.parquet 
 * https://d37ci6vzurychx.cloudfront.net/trip-data/green_tripdata_2023-10.parquet 
 * https://d37ci6vzurychx.cloudfront.net/trip-data/fhv_tripdata_2023-10.parquet
 * https://d37ci6vzurychx.cloudfront.net/trip-data/fhvhv_tripdata_2023-10.parquet 


```{r acquire-october-2023-tlc} 
## Aggregate all vehicular trips for October 2023
# Combine Yellow, Green, FHV, and High-Volume FHV (Uber/Lyft) 
# See "all" the car-based 'Last-Mile' choices

# URL definitions for source data
tlc_urls <- c(
  yellow = "https://d37ci6vzurychx.cloudfront.net/trip-data/yellow_tripdata_2023-10.parquet",
  green  = "https://d37ci6vzurychx.cloudfront.net/trip-data/green_tripdata_2023-10.parquet",
  #fhv    = "https://d37ci6vzurychx.cloudfront.net/trip-data/fhv_tripdata_2023-10.parquet",
  fhvhv  = "https://d37ci6vzurychx.cloudfront.net/trip-data/fhvhv_tripdata_2023-10.parquet"
)

# Function to read and count trips by Taxi Zone 
# PULocationID can be used to get the # of trips that started from (were picked up in) a zone
count_zone_trips <- function(url) { 
  read_parquet(url, col_select = any_of(c("PULocationID"))) %>%
    count(PULocationID, name = "trips") %>%
    filter(!is.na(PULocationID))
}

# Process the sources and sum them up 
# Getting number of trips that started per pick up location
# This 'map_dfr' loop downloads each, counts by zone, and stacks them.
total_zone_counts <- tlc_urls %>%
  map_dfr(count_zone_trips) %>%
  group_by(PULocationID) %>%
  summarize(total_vehicular_trips = sum(trips))
```

### Census uses Tracts, TLC uses Taxi Zones

So Census Tracts are not the same as Taxi Zones. This chunk tries to handle it.

**EPSG:2263** is apparently a specific map projection built just for the New York/Long Island area. In it, 1 unit equals exactly 1 physical foot. (I had not ever used this before.) The idea is that if the censues tracts and the taxi zones are both transformed to EPSG:2263, then this will be the most precise way to layer these maps on top of one another and see where they intersect. (At least it appears to be more precise than using latitude and longitude.)

`st_intersection()` lays the two maps on top of one another and separated out the overlapping pieces properly.

```{r} 
# Coordinate Reference System (CRS)
# EPSG:2263 (NAD83 / New York Long Island) uses feet 
nyc_tracts_proj <- st_transform(nyc_tracts, 2263)
taxi_zones_proj <- st_transform(taxi_zones, 2263)

# Intersect regions
# This creates a new spacial dataframe where tracts and zones overlap
intersection <- st_intersection(nyc_tracts_proj, taxi_zones_proj) %>%
  mutate(intersect_area = st_area(.))

# Calculate the weights
# For every unique Census Tract, 
# What % of Tract X's area is inside Taxi Zone Y?
tract_weights <- intersection %>%
  as_tibble() %>%
  group_by(GEOID) %>%
  mutate(total_tract_area = sum(intersect_area)) %>%
  ungroup() %>%
  mutate(weight = as.numeric(intersect_area / total_tract_area)) %>%
  select(GEOID, LocationID, zone, weight)

# Peak Crosswalk 
head(tract_weights)
```

The below chunk uses the newly defined geography, and then it disperses the uber/taxi trip data (which uses taxi zones) fairly across the census tracts.  
 
```{r}
# issue is that taxi zones are larger than census tracts, and we obvs want to use the data together 
# Assumes 'tract_weights' object exists from your earlier step.

tract_vehicular_estimates <- tract_weights %>%
  left_join(total_zone_counts, by = c("LocationID" = "PULocationID")) %>%
  mutate(estimated_trips = total_vehicular_trips * weight) %>%
  group_by(GEOID) %>%
  summarize(total_uber_taxi_trips = sum(estimated_trips, na.rm = TRUE))

# Save the 'Vehicular' side of the Master Table
save_path_tlc <- file.path(projPath, "data", "processed_tlc_counts.rds")
saveRDS(tract_vehicular_estimates, save_path_tlc)

print(paste("Vehicular trip estimates for October 2023 saved to:", save_path_tlc))
```
 
### Citibike

* [Citibike Data Download](https://s3.amazonaws.com/tripdata/index.html)
  * Because the Citibike publishes their data online in zip files, it can't be read directly from web.
* Originally was pulling **January 2026** data for citibike
  * 202601-citibike-tripdata_1.csv
  * 202601-citibike-tripdata_2.csv
* Currently pulling **October 2023** data  


```{r citi-bike-to-census-tract }
# Map Station IDs to Census Tracts  

# file locations
cb_file1 <- file.path(projPath, "citibike", "202310-citibike-tripdata_1.csv")
cb_file2 <- file.path(projPath, "citibike", "202310-citibike-tripdata_2.csv")
cb_file3 <- file.path(projPath, "citibike", "202310-citibike-tripdata_3.csv")
cb_file4 <- file.path(projPath, "citibike", "202310-citibike-tripdata_4.csv")
cb_files <- c(cb_file1, cb_file2, cb_file3, cb_file4)

# Station-to-Tract Lookup Table
# unique coordinates for every station found in files
all_stations <- cb_files %>%
  map_dfr(~read_csv(.x, 
                    col_select = c(start_station_id, start_lat, start_lng),
                    # Force the station ID to be a character, and coords to be numeric
                    col_types = cols(
                      start_station_id = col_character(),
                      start_lat = col_double(),
                      start_lng = col_double()
                    ))) %>%
  filter(!is.na(start_lat)) %>%
  distinct(start_station_id, .keep_all = TRUE)

# Spatial Work: Convert Stations to Points
stations_sf <- st_as_sf(all_stations, coords = c("start_lng", "start_lat"), crs = 4326) %>%
  st_transform(2263) # Project to NYC Feet (EPSG:2263)

# Ensure nyc_tracts (from Step 1) matches the projection
nyc_tracts_proj <- st_transform(nyc_tracts, 2263)

# Find which Tract Polygon physically contains each Station Point
station_tract_lookup <- st_join(stations_sf, nyc_tracts_proj, join = st_intersects) %>%
  as_tibble() %>%
  select(start_station_id, GEOID)

# Count the number of unique stations in each Census Tract
tract_station_counts <- station_tract_lookup %>%
  group_by(GEOID) %>%
  summarize(total_stations = n()) %>%
  filter(!is.na(GEOID))

# Count Trips per Tract 
tract_bike_counts <- cb_files %>%
  map_dfr(~read_csv(.x, 
                    col_select = c(start_station_id),
                    col_types = cols(start_station_id = col_character()) 
                    )) %>%
  left_join(station_tract_lookup, by = "start_station_id") %>%
  group_by(GEOID) %>%
  summarize(total_bike_trips = n()) %>%
  filter(!is.na(GEOID)) %>%
  # NEW: Attach the station capacity feature to the trip counts
  left_join(tract_station_counts, by = "GEOID")


# Save the final "Label" feature for the Machine Learning model
save_path <- file.path(projPath, "data", "processed_bike_counts.rds")
saveRDS(tract_bike_counts, save_path)

print(paste("Success! Bike counts saved to:", save_path))
```

### Distance from Subway

Source is [https://data.ny.gov/Transportation/MTA-Subway-Entrances-and-Exits-2024/i9wp-a4ja/about_data](https://data.ny.gov/Transportation/MTA-Subway-Entrances-and-Exits-2024/i9wp-a4ja/about_data). 

File name is "MTA_Subway_Entrances_and_Exits__2024.csv"

```{r subway-distance}
## Distance to Subway 

# Load csv
subway_csv_path <- file.path(projPath, "MTA_Subway_Entrances_and_Exits__2024.csv")
subway_raw <- read_csv(subway_csv_path) %>%
  clean_names() # otherwise this had spaces in the column names

# Transform into a Spatial Object (sf)
subway_entrances_sf <- subway_raw %>%
  filter(!is.na(entrance_latitude) & !is.na(entrance_longitude)) %>% # 'entrance_longitude' is X and 'entrance_latitude' is Y
  st_as_sf(coords = c("entrance_longitude", "entrance_latitude"), crs = 4326) %>% # telling it to read this as 4326, the standard GPS coordinate system  
  st_transform(2263) # Converting to the same 1 unit = 1 foot NYC-specific map that was used earlier

# Centroids
# Find the center point of every Census Tract
tract_centers <- st_centroid(nyc_tracts_proj)

# Nearest Entrance for every Tract
# Find the closest subway for each tract
# Doing this from the tract center 
nearest_indices <- st_nearest_feature(tract_centers, subway_entrances_sf)

# Distance in Miles
# (Distance in feet / 5280)
dist_to_subway_feet <- st_distance(tract_centers, subway_entrances_sf[nearest_indices, ], by_element = TRUE)

subway_dist_lookup <- nyc_tracts_proj %>%
  as_tibble() %>%
  mutate(dist_subway_miles = as.numeric(dist_to_subway_feet) / 5280) %>%
  select(GEOID, dist_subway_miles)

# Save the Subway Distance
saveRDS(subway_dist_lookup, file.path(projPath, "data", "subway_dist_lookup.rds"))

print("Subway distance feature successfully calculated from MTA data!")
```

## Assembling

### Big Join

```{r big-join}
## Big Join 

# Load components
census_features <- readRDS(file.path(projPath, "data", "nyc_census_tracts.rds"))
bike_counts     <- readRDS(file.path(projPath, "data", "processed_bike_counts.rds"))
tlc_counts      <- readRDS(file.path(projPath, "data", "processed_tlc_counts.rds"))
subway_dist     <- readRDS(file.path(projPath, "data", "subway_dist_lookup.rds"))

# Before joining, force each Tract to belong to only one Taxi Zone, the one it overlaps most.
# I was seeing repeating values of income and distance for different zones in the final data set back when I was not doing this. 
unique_weights <- tract_weights %>%
  group_by(GEOID) %>%
  slice_max(weight, n = 1, with_ties = FALSE) %>% 
  ungroup() %>%
  select(GEOID, LocationID, zone) # Keep the Zone name and ID for joining

# Join everything to the Census Tracts
# left_join to not lose any tracts that might have had 0 trips
master_table <- census_features %>%
  st_drop_geometry() %>% # drops the maps, keeps ALL features 
  # unique Neighborhood/LocationID mapping
  left_join(unique_weights, by = "GEOID") %>%
  # Bike counts (already at Tract level)
  left_join(bike_counts, by = "GEOID") %>%
  # Uber/Taxi counts  
  left_join(tlc_counts, by = "GEOID") %>% 
  # Subway Distance (already at Tract level)
  left_join(subway_dist, by = "GEOID")


# Clean up NAs; create the Target Variable
# If a tract has no recorded trips, replace NA with 0
# If a tract has no stations, replace NA with 0 
master_table <- master_table %>%
  mutate(
    total_bike_trips      = replace_na(total_bike_trips, 0),
    total_uber_taxi_trips = replace_na(total_uber_taxi_trips, 0),
    total_stations        = replace_na(total_stations, 0),
    
    # Spatial Density (Stations per Square Mile)
    # How physically clustered is the infrastructure?
    stations_per_sq_mile  = total_stations / area_sq_miles,
    
    # Per Capita Availability (Stations per 1,000 Residents)
    # How much infrastructure exists relative to the local population size?
    stations_per_1000_pop = (total_stations / total_pop) * 1000,
    uber_trips_per_capita = total_uber_taxi_trips / total_pop,
    bike_trips_per_capita = total_bike_trips / total_pop,
    
    # Old target: % of Last-Mile trips taken by Bike
    # I left this in here, but we aren't really using it now
    bike_index = total_bike_trips / (total_bike_trips + total_uber_taxi_trips)
  )  

# Save joined data set
saveRDS(master_table, file.path(projPath, "data", "model_input.rds"))

print("Master Join Complete. The model_input.rds file is ready for modeling.")
```



### Intermediate Sanity Checking

We could drop this later, but right now it was just a helpful way to make sure that combining the census tracts and the taxi zones worked properly.

```{r data-audit}
### The Data Audit (Sanity Check)

# Load the Master Table
master_table <- readRDS(file.path(projPath, "data", "model_input.rds"))

# Join with Taxi Zone names for readability
# We'll use the tract_weights crosswalk to get neighborhood names back
audit_table <- master_table %>%
  select(zone, GEOID, bike_index, total_bike_trips, total_uber_taxi_trips, med_income, dist_subway_miles)

# Top 10 "Bike-Heavy" Neighborhoods (Active)
top_bike <- audit_table %>%
  arrange(desc(bike_index)) %>%
  head(50)

# Top 10 "Uber-Heavy" Neighborhoods (Vehicular)
top_uber <- audit_table %>%
  arrange(bike_index) %>%
  head(50)

# quick summary
print("--- TOP 10 BIKE-HEAVY TRACTS ---")
print(top_bike)

print("--- TOP 10 UBER-HEAVY TRACTS ---")
print(top_uber)
```


## Model Building (Super Duper Rough, Proof of Concept)

This is currently a simple linear model for proof of concept. We'll be able to build up something much more robust .

### Borough Fixed Effects

Our professor flagged that NYC neighborhoods are very different from each other — a feature value in the Bronx means something different than the same value in Manhattan. Adding borough as a factor variable lets both models learn a borough-level intercept, capturing unobserved variation that our census features don't fully explain.

Borough is derived from the first 5 digits of GEOID (standard Census FIPS codes) so no additional data pull is needed.

```{r borough-feature}
# FIPS county codes for NYC boroughs
borough_lookup <- c(
  "36005" = "Bronx",
  "36047" = "Brooklyn",
  "36061" = "Manhattan",
  "36081" = "Queens",
  "36085" = "Staten Island"
)

# Derive borough from first 5 chars of GEOID and join onto the master table
master_table <- readRDS(file.path(projPath, "data", "model_input.rds")) %>%
  mutate(borough = factor(borough_lookup[substr(GEOID, 1, 5)]))

saveRDS(master_table, file.path(projPath, "data", "model_input.rds"))
print(table(master_table$borough))
```

### Model 1

`ml_data` is selecting the features to go into the model. A very very simple model is happening here. We need to implement train-test and better model selection still.

```{r machine-learning-poc}
# The selection for the Machine Learning model
# Right now, this has a lot of room for improvement trimming or adding features 

# To make iteration easier:
##  COLUMN/FEATURE SELECTION
id_vars <- c("GEOID")
target_var <- c("pct_commute_60p")

predictor_vars <- c(
  # Borough Fixed Effect
  "borough",

  # Infrastructure & Mobility
  "stations_per_sq_mile", "stations_per_1000_pop", "dist_subway_miles",
  "uber_trips_per_capita", "bike_trips_per_capita",

  # Economic & Housing
  "med_income", "pct_poverty", "pct_vacant", "pct_renter",

  # Demographics & Commute Behaviors
  "pop_density", "pct_no_car", "pct_transit", "pct_bachelors", "pct_wfh", "pct_walk",

  # Age Buckets
  "pct_age_18_24", "pct_age_25_34", "pct_age_35_44", "pct_age_45_64", "pct_age_65_plus"
)

## SAVING RAW (NON-Z-SCORE) DATA TO DISPLAY IN SHINY APP
ml_data_raw <- readRDS(file.path(projPath, "data", "model_input.rds")) %>%
  # all_of() safely selects only the columns defined in our lists above
  select(all_of(c(id_vars, target_var, predictor_vars))) %>%
  drop_na()

## SCALE THE DATA FOR THE MODEL
# Scale numeric predictors only — borough is a factor and must not be z-scored
numeric_predictors <- predictor_vars[predictor_vars != "borough"]

ml_data_scaled <- ml_data_raw %>%
  mutate(across(
    .cols = all_of(numeric_predictors),
    .fns  = ~ as.numeric(scale(.))
  ))

## TRAIN THE MODEL
# Train the model using the SCALED data
lm_fit_standardized <- lm(pct_commute_60p ~ . - GEOID, data = ml_data_scaled)

print("Model Summary: ")
summary(lm_fit_standardized) 
```

### Model 2: Random Forest + Evaluation

Adding a Random Forest alongside the linear baseline. Both are evaluated on a held-out test set so the comparison is fair. The RF is also trained on the full dataset (`rf_fit`) for use in the IBX simulation and map predictions below.

```{r random-forest-model}
library(randomForest)

## TRAIN/TEST SPLIT (80/20)
set.seed(622)
train_idx  <- sample(nrow(ml_data_scaled), size = 0.8 * nrow(ml_data_scaled))
train_data <- ml_data_scaled[train_idx, ]
test_data  <- ml_data_scaled[-train_idx, ]

## Retrain linear on train set only (for a fair apples-to-apples comparison)
lm_fit_eval <- lm(pct_commute_60p ~ ., data = train_data %>% select(-GEOID))

## Train RF on train set (for evaluation)
# Drop GEOID from data explicitly — randomForest treats character columns as
# factors and fails on new levels at predict time if GEOID is left in
rf_fit_eval <- randomForest(
  pct_commute_60p ~ .,
  data       = train_data %>% select(-GEOID),
  ntree      = 500,
  importance = TRUE
)

## Train RF on full dataset (used for map predictions)
rf_fit <- randomForest(
  pct_commute_60p ~ .,
  data       = ml_data_scaled %>% select(-GEOID),
  ntree      = 500,
  importance = TRUE
)

## EVALUATION on test set
lm_preds <- predict(lm_fit_eval, newdata = test_data %>% select(-GEOID))
rf_preds <- predict(rf_fit_eval, newdata = test_data %>% select(-GEOID))
actuals  <- test_data$pct_commute_60p

rmse <- function(a, p) sqrt(mean((a - p)^2))
r2   <- function(a, p) 1 - sum((a - p)^2) / sum((a - mean(a))^2)

eval_table <- tibble(
  Model = c("Linear Regression", "Random Forest"),
  RMSE  = round(c(rmse(actuals, lm_preds), rmse(actuals, rf_preds)), 3),
  R2    = round(c(r2(actuals,   lm_preds), r2(actuals,   rf_preds)), 3)
)

kable(eval_table, caption = "Model Comparison — 80/20 Test Set")
```

### IBX Simulation

Here, the approximate locations of the new IBX stations are being brought in. And then in the census areas where they go, we're basically dropping the distance to train for those areas to really low. There may be a more robust way of implementing the "simulation" of the IBX station than this, but this seemed like a good starting point.

```{r ibx-simulator-data} 

# Build the list of stations based on the MTA's 2025 CE Document
# This document is a pdf, and can be found in the repo
ibx_stations <- tibble::tribble(
  ~station_name,             ~address,
  "Brooklyn Army Terminal",  "2nd Ave & 61st St, Brooklyn, NY",
  "4th Ave",                 "5th Ave & 61st St, Brooklyn, NY",
  "8th Ave",                 "8th Ave & 61st St, Brooklyn, NY",
  "New Utrecht Ave",         "14th Ave & 61st St, Brooklyn, NY",
  "McDonald Ave",            "McDonald Ave & Avenue I, Brooklyn, NY",
  "East 16th St",            "East 16th St & Avenue H, Brooklyn, NY",
  "Flatbush-Nostrand Ave",   "Nostrand Ave & Avenue H, Brooklyn, NY",
  "Utica Ave",               "Utica Ave & Farragut Rd, Brooklyn, NY",
  "Remsen Ave",              "Remsen Ave & Farragut Rd, Brooklyn, NY",
  "Linden Blvd",             "Linden Blvd & Junius St, Brooklyn, NY",
  "Livonia Ave",             "Livonia Ave & Junius St, Brooklyn, NY",
  "Sutter Ave",              "Sutter Ave & Junius St, Brooklyn, NY",
  "Atlantic Ave",            "Atlantic Ave & East New York Ave, Brooklyn, NY",
  "Wilson Ave",              "Wilson Ave & Moffat St, Brooklyn, NY",
  "Myrtle Ave",              "Cypress Hills St & 78th Ave, Queens, NY",
  "Metropolitan Ave",        "Metropolitan Ave & 69th St, Queens, NY",
  "Eliot Ave",               "Eliot Ave & 69th St, Queens, NY",
  "Grand Ave",               "Grand Ave & 69th St, Queens, NY",
  "Roosevelt Ave",           "Roosevelt Ave & 74th St, Queens, NY"
)

# Geocode the addresses into Lat/Long coordinates
ibx_geocoded <- ibx_stations %>%
  geocode(address, method = 'osm', lat = latitude, long = longitude) %>% # get coordinated for each address
  drop_na() %>% # Drop any that the open-source map couldn't perfectly match
  
  # Convert the GPS coordinates into an sf object
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  
  # Convert and draw 10-minute-walk-radiuses 
  st_transform(2263) %>% # transform to the more precide NYC + Long Island map we've been using
  st_buffer(dist = 2640) %>% # drawing a circle around point that is 2640 feet = 0.5 miles ~ 10 minute walk. Most people don't want to walk further than 10 minutes to the train
  st_transform(4326) # Convert back to standard GPS

# Find all Census Tracts that fall at least partially inside these new 10-minute-walk-radiuses 
ibx_tracts_sf <- st_filter(
  nyc_tracts, 
  st_transform(ibx_geocoded, st_crs(nyc_tracts)), 
  .predicate = st_intersects
)

# Extract just the 11-digit IDs as a clean list
ibx_geoid_list <- ibx_tracts_sf$GEOID
```

```{r ibx-simulator-dataset-and-model}
## Create an IBX UNIVERSE data set
# finding the Census Tract in NYC that is physically closest to a subway entrance
# going to use this later to simulate a really close distance to subway for places where IBX is being built
best_transit_z_score <- min(ml_data_scaled$dist_subway_miles)

ibx_scenario_data_scaled <- ml_data_scaled %>%
  mutate(
     # Inject best_transit_z_score into the IBX tracts
    dist_subway_miles = ifelse(GEOID %in% ibx_geoid_list, best_transit_z_score, dist_subway_miles)
  )

# Predicted data is from the model which consumed scaled data
# But we're using raw data so the display of neighborhood features is correct
final_map_data <- ml_data_raw %>%
  mutate(
    predicted_commute_baseline = predict(lm_fit_standardized, newdata = ml_data_scaled),
    predicted_commute_ibx      = predict(lm_fit_standardized, newdata = ibx_scenario_data_scaled),
    ibx_commute_relief         = predicted_commute_baseline - predicted_commute_ibx,
    is_ibx_location            = ifelse(GEOID %in% ibx_geoid_list, 1, 0)
  )

saveRDS(final_map_data, file.path(projPath, "data", "model_input.rds")) 
```

```{r ibx-rf-predictions}
# Append RF IBX predictions and residuals to the map data already saved by the chunk above.
# final_map_data is still in memory — adding columns and re-saving.
final_map_data_rf <- final_map_data %>%
  mutate(
    predicted_commute_baseline_rf = pmax(0, predict(rf_fit, newdata = ml_data_scaled %>% select(-GEOID))),
    predicted_commute_ibx_rf      = pmax(0, predict(rf_fit, newdata = ibx_scenario_data_scaled %>% select(-GEOID))),
    ibx_commute_relief_rf         = pmax(0, predicted_commute_baseline_rf - predicted_commute_ibx_rf),
    # Residual: positive = model underestimated commute burden (worse than expected)
    #           negative = model overestimated (better than expected)
    rf_residual                   = pct_commute_60p - predicted_commute_baseline_rf
  )

saveRDS(final_map_data_rf, file.path(projPath, "data", "model_input.rds"))
print("RF predictions and residuals appended to model_input.rds")
```

```{r model-outputs}

# Model Results
lm_results <- tidy(lm_fit_standardized) %>%
  filter(term != "(Intercept)") # Remove the intercept for better scaling

# Plot the "Pull" or "Impact" of each feature
ggplot(lm_results, aes(x = estimate, y = reorder(term, estimate))) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  geom_point(size = 3, color = "darkblue") +
  geom_errorbarh(aes(xmin = estimate - std.error, xmax = estimate + std.error), height = 0.2) +
  theme_minimal() +
  labs(title = "What's driving 60 minute + length communtes?",
       subtitle = "Standardized Coefficients (Variables further from 0 have a stronger impact)",
       x = "Effect on 60+ Minute Commute Rate (per 1 Std. Dev. increase)",
       y = "Feature")
```

```{r rf-feature-importance}
# Variable importance from the full RF model
rf_importance <- importance(rf_fit) %>%
  as.data.frame() %>%
  rownames_to_column("Feature") %>%
  arrange(desc(`%IncMSE`))

ggplot(rf_importance, aes(x = `%IncMSE`, y = reorder(Feature, `%IncMSE`))) +
  geom_point(size = 3, color = "darkgreen") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(
    title    = "Random Forest: Feature Importance",
    subtitle = "% increase in MSE when feature is permuted — higher = more important",
    x        = "% Increase in MSE (Permutation Importance)",
    y        = "Feature"
  )
```

## Misc. Notes Dump
 
nothing right now.