---
title: "DATA 622 Project"
author: "Catherine Dube, Guillermo Schneider, Tai Chou-Kudu"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  out.width = "100%", 
  fig.align = "center"
)

if(!require(pacman)) install.packages("pacman")
pacman::p_load(
   tidyverse, 
   fpp3, 
   rstudioapi,
   tidycensus,
   r5r, 
   sf, #spacial data handling
   tigris, #NYC geography
   mapview,
   osmextract,
   arrow, # read parquet data, designed for random access?
   dplyr,
   knitr,
   janitor # clean up the names of columns that have spaces (subway entrances data)
)
projPath <- dirname(file.path(getSourceEditorContext()$path))

```


# Modeling Citi Bike vs. Rideshare Usage in NYC

## Foundation 

### Census Data 

Looking at census data from **2022**. 

Picked 2022 because:

* This is a good Post-Pandemic option; NYC was "fully reopened" 
* The 5-Year ACS helps avoid the 2020 collection gap (where due to COVID, the Census Bureau couldn't send people door to door) and the 2021 remote-forward living arrangements, where many people left Manhattan during the high period of remote work.
* The 5-year estimate is a rolling average (2018–2022), and some COVID-impacted data is included either way. However, multiple years of more robust data are included in this time window.

Why is this data important?

* The census data provides **Feature variables** that describe what New Yorkers looked like in a given region. These are inputs. 

**REVISIT ME:** 

* Currently: looking at vehicle availability, household income, age, and population size.
* Future: probably want to comb through other features of interest from the census data. Potentially look at feature engineering, and at normalizing or standardizing the features as needed.

```{r acquire-census-data} 

#CENSUS_API_KEY is in environment 

nyc_counties <- c("New York", "Kings", "Queens", "Bronx", "Richmond")

 
# Entire 2022 5-Year ACS variable dictionary
acs_vars_2022 <- load_variables(2022, "acs5", cache = TRUE)

# MISC.
# Exploring different potential features
feature_exploration <- acs_vars_2022 %>%
  filter(
    str_detect(str_to_lower(label), "public transportation|bicycle|walked|worked from home") |
    str_detect(str_to_lower(concept), "means of transportation to work|travel time")
  )

# Comment Me out before final knit
#View(feature_exploration)
```

```{r more-census-data}

# Define all API variables explicitly
vars <- c(
  # Base Demographics
  no_car      = "B08201_002", 
  med_income  = "B19013_001", 
  med_age     = "B01002_001",
  total_pop   = "B01003_001",
  
  # Transportation Base & Features
  tot_workers = "B08301_001",
  transit     = "B08301_010",
  walk        = "B08301_019",
  wfh         = "B08301_021",
  
  # Travel Time Features
  commute_60  = "B08303_012",
  commute_90  = "B08303_013",
  
  # Education Base & Features
  tot_pop_25  = "B15003_001",
  bachelors   = "B15003_022",
  
  # Occupation Base & The 5 Major Buckets
  tot_employ  = "C24010_001",
  mngmt_job   = "C24010_002",
  service_job = "C24010_014",
  sales_job   = "C24010_021",
  const_job   = "C24010_030",
  trans_job   = "C24010_034"
)

# Getting from census data
nyc_tracts <- get_acs(
  geography = "tract",
  variables = vars,
  state = "NY",
  county = nyc_counties,
  geometry = TRUE, 
  year = 2022,
  output = "wide"
)

```

```{r feature-engineering-census-data}
nyc_tracts <- nyc_tracts %>%
  select(
    GEOID, 
    NAME, 
    geometry,
    no_car      = no_carE,
    med_income  = med_incomeE,
    med_age     = med_ageE,
    total_pop   = total_popE,
    tot_workers = tot_workersE,
    tot_pop_25  = tot_pop_25E,
    tot_employ  = tot_employE,
    transit     = transitE,
    walk        = walkE,
    wfh         = wfhE,
    commute_60  = commute_60E,
    commute_90  = commute_90E,
    bachelors   = bachelorsE,
    mngmt_job   = mngmt_jobE,
    service_job = service_jobE,
    sales_job   = sales_jobE,
    const_job   = const_jobE,
    trans_job   = trans_jobE
  ) %>%
  
  # Feature Engineering: Convert raw counts to percentages
  mutate(
    pct_no_car      = (no_car / total_pop) * 100,
    pct_transit     = (transit / tot_workers) * 100,
    pct_walk        = (walk / tot_workers) * 100,
    pct_wfh         = (wfh / tot_workers) * 100,
    
    # Percentage of Commuters Traveling 60+ Minutes
    pct_commute_60p = ((commute_60 + commute_90) / tot_workers) * 100,
    
    pct_bachelors   = (bachelors / tot_pop_25) * 100,
    
    # Occupation Percentages 
    pct_mngmt       = (mngmt_job / tot_employ) * 100,
    pct_service     = (service_job / tot_employ) * 100,
    pct_sales       = (sales_job / tot_employ) * 100,
    pct_const       = (const_job / tot_employ) * 100,
    pct_trans       = (trans_job / tot_employ) * 100,
    
    # Population Density (people per square mile)
    area_sq_miles   = as.numeric(st_area(geometry)) / 2589988, # st_area() returns square meters, converted
    pop_density     = total_pop / area_sq_miles
  ) %>%
  
  # Drop the raw counts because I don't want to accidentally put them in the ML model
  select(-no_car, -transit, -walk, -wfh, -commute_60, -commute_90, -bachelors, 
         -mngmt_job, -service_job, -sales_job, -const_job, -trans_job, 
         -tot_workers, -tot_pop_25, -tot_employ)
```

```{r census-missing-data-exploration}
## Checking for total NAs and NaNs in each column
colSums(is.na(nyc_tracts %>% st_drop_geometry())) %>%
  kable(caption = "# of NAs and NaNs in each column")

## Percentage of missing data 
# (is it a low enough percent to be negligible) 
missing_summary <- nyc_tracts %>%
  st_drop_geometry() %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(cols = everything(), names_to = "Feature", values_to = "Missing_Count") %>%
  mutate(
    Total_Rows = nrow(nyc_tracts),
    Pct_Missing = (Missing_Count / Total_Rows) * 100,
    
    # Flag if the missing data is below 5% 
    Is_Negligible_Under_5Pct = Pct_Missing < 5 
  ) %>%
  # Filter to only show features that actually have missing data
  filter(Missing_Count > 0) %>% 
  arrange(desc(Pct_Missing))

# Output Percentage of Missing Data
kable(missing_summary, digits = 2, caption = "Missing Data Percentage by Feature")


## Map for missing income data
missing_map_data <- nyc_tracts %>%
  mutate(missing_income_flag = ifelse(is.na(med_income), "Missing Data", "Valid Data"))

ggplot(missing_map_data) +
  geom_sf(aes(fill = missing_income_flag), color = "white", size = 0.1) +
  scale_fill_manual(values = c("Valid Data" = "#e0e0e0", "Missing Data" = "#d73027")) +
  theme_void() +
  labs(title = "Map of Missing Census Income Data",
       fill = "Status") 
```

```{r census-na-handling}
## Drop tracts with less than 50 people or 0 workers
# There were a ton of tracts with a small number of or 0 people, screwing with denominators and also making for super volatile percentages
nyc_tracts_clean <- nyc_tracts %>%
  filter(total_pop >= 50, !is.na(pct_transit))

## Imputing
# There were also many tracts where income or age was missing, but it wasn't due to a denominator issue
## RETURN HERE
## Just doing simple median imputation
## More robust method would be better
## Checking for MNAR and stuff would be better
# Right now
nyc_tracts_clean <- nyc_tracts_clean %>%
  # Extract the borough from the NAME string (e.g., "Queens County")
  mutate(borough = str_extract(NAME, "(?<=; ).*(?= County)")) %>%
  
  # Group by Borough so the imputation is localized
  group_by(borough) %>%
  mutate(
    med_income = ifelse(is.na(med_income), 
                        median(med_income, na.rm = TRUE), 
                        med_income),
    med_age = ifelse(is.na(med_age), 
                     median(med_age, na.rm = TRUE), 
                     med_age)
  ) %>%
  ungroup() %>%
  
  # Drop the temporary borough column 
  select(-borough)
 
print(paste("Original tracts:", nrow(nyc_tracts)))
print(paste("Cleaned tracts of NAs:", nrow(nyc_tracts_clean)))
```


```{r save-census-data}
# Save 
census_save_path <- file.path(projPath, "data", "nyc_census_tracts.rds") 
saveRDS(nyc_tracts_clean, census_save_path)

print(paste("Census features and geometry saved to:", census_save_path))

# Peak
head(nyc_tracts_clean)
```

### Taxi Zones

```{r taxi-zones}  
#Zip file from below url is downloaded and extracted into repo
#taxi_zones_url <- "https://d37ci6vzurychx.cloudfront.net/misc/taxi_zones.zip" 

# Read the shapefile
taxi_zones_sf <- st_read(file.path(projPath, 'taxi_zones', 'taxi_zones', "taxi_zones.shp")) %>%
  st_transform(4326) # Transform to standard Lat/Long (WGS84)

# Lookup table with names and boroughs is downloaded into repo
#zone_lookup_url <- "https://d37ci6vzurychx.cloudfront.net/misc/taxi_zone_lookup.csv"
zone_lookup <- read_csv(file.path(projPath, 'taxi_zone_lookup.csv'))

# Join 
taxi_zones <- taxi_zones_sf %>%
  left_join(zone_lookup, by = c("LocationID" = "LocationID"))

# Peak: Does this map look right 
plot(st_geometry(taxi_zones))
```

### TLC Data

TLC Data is from the [NYC Taxi and Limousine Commission](https://www.nyc.gov/site/tlc/about/tlc-trip-record-data.page).

High volume for hire vehicles = Uber/Lyft.

* Originally was pulling **January 2026** data for High Volume for-hire vehicles.
 * https://d37ci6vzurychx.cloudfront.net/trip-data/fhvhv_tripdata_2026-01.parquet 
* Currently pulling **October 2023** data for yellow taxis, green taxis, ~for-hire vehicles (FHV (local livery/black car)),~ and high volume for-hire vehicles (FHVHV (Uber/Lyft) Uber/Lyft). This made a little more sense with the 2022 census data, and with the seasonality of commuting and to avoid holiday season. 
 * https://d37ci6vzurychx.cloudfront.net/trip-data/yellow_tripdata_2023-10.parquet 
 * https://d37ci6vzurychx.cloudfront.net/trip-data/green_tripdata_2023-10.parquet 
 * https://d37ci6vzurychx.cloudfront.net/trip-data/fhv_tripdata_2023-10.parquet
 * https://d37ci6vzurychx.cloudfront.net/trip-data/fhvhv_tripdata_2023-10.parquet 

**REVISIT ME:** 

* Currently:   
* Future:  
 
```{r acquire-october-2023-tlc} 
## Aggregate all vehicular trips for October 2023
# Combine Yellow, Green, FHV, and High-Volume FHV (Uber/Lyft) 
# See "all" the car-based 'Last-Mile' choices

# URL definitions for source data
tlc_urls <- c(
  yellow = "https://d37ci6vzurychx.cloudfront.net/trip-data/yellow_tripdata_2023-10.parquet",
  green  = "https://d37ci6vzurychx.cloudfront.net/trip-data/green_tripdata_2023-10.parquet",
  #fhv    = "https://d37ci6vzurychx.cloudfront.net/trip-data/fhv_tripdata_2023-10.parquet",
  fhvhv  = "https://d37ci6vzurychx.cloudfront.net/trip-data/fhvhv_tripdata_2023-10.parquet"
)

# Function to read and count trips by Taxi Zone 
# PULocationID can be used to get the # of trips that started from (were picked up in) a zone
count_zone_trips <- function(url) { 
  read_parquet(url, col_select = any_of(c("PULocationID"))) %>%
    count(PULocationID, name = "trips") %>%
    filter(!is.na(PULocationID))
}

# Process the sources and sum them up 
# Getting number of trips that started per pick up location
# This 'map_dfr' loop downloads each, counts by zone, and stacks them.
total_zone_counts <- tlc_urls %>%
  map_dfr(count_zone_trips) %>%
  group_by(PULocationID) %>%
  summarize(total_vehicular_trips = sum(trips))
```

### Census uses Tracts, TLC uses Taxi Zones

The question here is how much of each Census Tract sits inside which Taxi Zone?

This uses EPSG:2263 which uses the units of "feet", and can be used to calculate the square footage of an area in NY more precisely.

```{r} 
# Use Coordinate Reference System (CRS)
# We'll use EPSG:2263 (NAD83 / New York Long Island) because it uses feet,
# which is much better for calculating area in NYC than Lat/Long.
nyc_tracts_proj <- st_transform(nyc_tracts, 2263)
taxi_zones_proj <- st_transform(taxi_zones, 2263)

# Intersect regions
# This creates a new set of shapes where tracts and zones overlap
intersection <- st_intersection(nyc_tracts_proj, taxi_zones_proj) %>%
  mutate(intersect_area = st_area(.))

# Calculate the weights
# For every unique Census Tract, 
# What % of Tract X's area is inside Taxi Zone Y?
tract_weights <- intersection %>%
  as_tibble() %>%
  group_by(GEOID) %>%
  mutate(total_tract_area = sum(intersect_area)) %>%
  ungroup() %>%
  mutate(weight = as.numeric(intersect_area / total_tract_area)) %>%
  select(GEOID, LocationID, zone, weight)

# Peak Crosswalk 
head(tract_weights)
```

 
```{r}

# 'Spatial Weights' / Crosswalk
# issue is that taxi zones are larger than census tracts, and we obvs want to use the data together
# 
# WHY: Taxi Zones are larger than Census Tracts. We use the weights from Phase 1
# to distribute these trips into our Census Tracts (GEOIDs).
# Assumes 'tract_weights' object exists from your earlier step.

tract_vehicular_estimates <- tract_weights %>%
  left_join(total_zone_counts, by = c("LocationID" = "PULocationID")) %>%
  mutate(estimated_trips = total_vehicular_trips * weight) %>%
  group_by(GEOID) %>%
  summarize(total_uber_taxi_trips = sum(estimated_trips, na.rm = TRUE))

# Save the 'Vehicular' side of the Master Table
save_path_tlc <- file.path(projPath, "data", "processed_tlc_counts.rds")
saveRDS(tract_vehicular_estimates, save_path_tlc)

print(paste("Vehicular trip estimates for October 2023 saved to:", save_path_tlc))
```
 
### Citibike

* [Citibike Data Download](https://s3.amazonaws.com/tripdata/index.html)
  * Because the Citibike publishes their data online in zip files, it can't be read directly from web.
* Originally was pulling **January 2026** data for citibike
  * 202601-citibike-tripdata_1.csv
  * 202601-citibike-tripdata_2.csv
* Currently pulling **October 2023** data  
  * 


```{r citi-bike-to-census-tract }
# Map Station IDs to Census Tracts  

# file locations
cb_file1 <- file.path(projPath, "citibike", "202310-citibike-tripdata_1.csv")
cb_file2 <- file.path(projPath, "citibike", "202310-citibike-tripdata_2.csv")
cb_file3 <- file.path(projPath, "citibike", "202310-citibike-tripdata_3.csv")
cb_file4 <- file.path(projPath, "citibike", "202310-citibike-tripdata_4.csv")
cb_files <- c(cb_file1, cb_file2, cb_file3, cb_file4)

# Station-to-Tract Lookup Table
# unique coordinates for every station found in files
all_stations <- cb_files %>%
  map_dfr(~read_csv(.x, 
                    col_select = c(start_station_id, start_lat, start_lng),
                    # Force the station ID to be a character, and coords to be numeric
                    col_types = cols(
                      start_station_id = col_character(),
                      start_lat = col_double(),
                      start_lng = col_double()
                    ))) %>%
  filter(!is.na(start_lat)) %>%
  distinct(start_station_id, .keep_all = TRUE)

# Spatial Work: Convert Stations to Points
stations_sf <- st_as_sf(all_stations, coords = c("start_lng", "start_lat"), crs = 4326) %>%
  st_transform(2263) # Project to NYC Feet (EPSG:2263)

# Ensure nyc_tracts (from Step 1) matches the projection
nyc_tracts_proj <- st_transform(nyc_tracts, 2263)

# Find which Tract Polygon physically contains each Station Point
station_tract_lookup <- st_join(stations_sf, nyc_tracts_proj, join = st_intersects) %>%
  as_tibble() %>%
  select(start_station_id, GEOID)

# Count the number of unique stations in each Census Tract
tract_station_counts <- station_tract_lookup %>%
  group_by(GEOID) %>%
  summarize(total_stations = n()) %>%
  filter(!is.na(GEOID))

# Count Trips per Tract 
tract_bike_counts <- cb_files %>%
  map_dfr(~read_csv(.x, 
                    col_select = c(start_station_id),
                    col_types = cols(start_station_id = col_character()) 
                    )) %>%
  left_join(station_tract_lookup, by = "start_station_id") %>%
  group_by(GEOID) %>%
  summarize(total_bike_trips = n()) %>%
  filter(!is.na(GEOID)) %>%
  # NEW: Attach the station capacity feature to the trip counts
  left_join(tract_station_counts, by = "GEOID")


# Save the final "Label" feature for the Machine Learning model
save_path <- file.path(projPath, "data", "processed_bike_counts.rds")
saveRDS(tract_bike_counts, save_path)

print(paste("Success! Bike counts saved to:", save_path))
```

### Distance from Subway

Source is [https://data.ny.gov/Transportation/MTA-Subway-Entrances-and-Exits-2024/i9wp-a4ja/about_data](https://data.ny.gov/Transportation/MTA-Subway-Entrances-and-Exits-2024/i9wp-a4ja/about_data). 

File name is "MTA_Subway_Entrances_and_Exits__2024.csv"

```{r subway-distance}
## Distance to Subway 

# Load csv
subway_csv_path <- file.path(projPath, "MTA_Subway_Entrances_and_Exits__2024.csv")
subway_raw <- read_csv(subway_csv_path) %>%
  clean_names() # otherwise this had spaces in the column names

# Transform into a Spatial Object (sf)
# 'entrance_longitude' is X and 'entrance_latitude' is Y
# 4326 is the standard GPS system (WGS84)
subway_entrances_sf <- subway_raw %>%
  filter(!is.na(entrance_latitude) & !is.na(entrance_longitude)) %>%
  st_as_sf(coords = c("entrance_longitude", "entrance_latitude"), crs = 4326) %>%
  st_transform(2263) # Convert to NYC Feet to match our Census Tracts

# Centroids
# Find the center point of every Census Tract
tract_centers <- st_centroid(nyc_tracts_proj)

# Nearest Entrance for every Tract
# Find the closest subway for each tract
nearest_indices <- st_nearest_feature(tract_centers, subway_entrances_sf)

# Distance in Miles
# (Distance in feet / 5280)
dist_to_subway_feet <- st_distance(tract_centers, subway_entrances_sf[nearest_indices, ], by_element = TRUE)

subway_dist_lookup <- nyc_tracts_proj %>%
  as_tibble() %>%
  mutate(dist_subway_miles = as.numeric(dist_to_subway_feet) / 5280) %>%
  select(GEOID, dist_subway_miles)

# Save the Subway Distance
saveRDS(subway_dist_lookup, file.path(projPath, "data", "subway_dist_lookup.rds"))

print("Subway distance feature successfully calculated from MTA data!")
```

## Assembling

### Big Join

```{r big-join}
## Big Join 

# Load components
census_features <- readRDS(file.path(projPath, "data", "nyc_census_tracts.rds"))
bike_counts     <- readRDS(file.path(projPath, "data", "processed_bike_counts.rds"))
tlc_counts      <- readRDS(file.path(projPath, "data", "processed_tlc_counts.rds"))
subway_dist     <- readRDS(file.path(projPath, "data", "subway_dist_lookup.rds"))

# Before joining, force each Tract to belong to only one Taxi Zone, the one it overlaps most.
# I was seeing repeating values of income and distance for different zones in the final data set back when I was not doing this. 
unique_weights <- tract_weights %>%
  group_by(GEOID) %>%
  slice_max(weight, n = 1, with_ties = FALSE) %>% 
  ungroup() %>%
  select(GEOID, LocationID, zone) # Keep the Zone name and ID for joining

# Join everything to the Census Tracts
# left_join to not lose any tracts that might have had 0 trips
master_table <- census_features %>%
  st_drop_geometry() %>% # drops the maps, keeps ALL features 
  # unique Neighborhood/LocationID mapping
  left_join(unique_weights, by = "GEOID") %>%
  # Bike counts (already at Tract level)
  left_join(bike_counts, by = "GEOID") %>%
  # Uber/Taxi counts  
  left_join(tlc_counts, by = "GEOID") %>% 
  # Subway Distance (already at Tract level)
  left_join(subway_dist, by = "GEOID")


# Clean up NAs; create the Target Variable
# If a tract has no recorded trips, replace NA with 0
# If a tract has no stations, replace NA with 0
master_table <- master_table %>%
  mutate(
    total_bike_trips      = replace_na(total_bike_trips, 0),
    total_uber_taxi_trips = replace_na(total_uber_taxi_trips, 0),
    total_stations        = replace_na(total_stations, 0), 
    
    # CALCULATE THE TARGET: % of Last-Mile trips taken by Bike
    bike_index = total_bike_trips / (total_bike_trips + total_uber_taxi_trips)
  ) %>%
  filter(!is.na(bike_index))

# Save joined data set
saveRDS(master_table, file.path(projPath, "data", "model_input.rds"))

print("Master Join Complete! The model_input.rds file is ready for Model Phase.")
```

### Intermediate Sanity Checking



```{r}
### Phase 2.5: The Data Audit (Sanity Check)

# 1. Load the Master Table we just created
master_table <- readRDS(file.path(projPath, "data", "model_input.rds"))

# 2. Join with Taxi Zone names for readability
# We'll use the tract_weights crosswalk to get neighborhood names back
audit_table <- master_table %>%
  select(zone, GEOID, bike_index, total_bike_trips, total_uber_taxi_trips, med_income, dist_subway_miles)

# 3. Top 10 "Bike-Heavy" Neighborhoods (Active)
top_bike <- audit_table %>%
  arrange(desc(bike_index)) %>%
  head(50)

# 4. Top 10 "Uber-Heavy" Neighborhoods (Vehicular)
top_uber <- audit_table %>%
  arrange(bike_index) %>%
  head(50)

# 5. Print a quick summary
print("--- TOP 10 BIKE-HEAVY TRACTS ---")
print(top_bike)

print("--- TOP 10 UBER-HEAVY TRACTS ---")
print(top_uber)
```


## Model Building (Super Duper Rough, Proof of Concept)

It's late in the evening, so I just did a quick AI pass at this with a linear model for proof of concept.

###

```{r machine-learning-poc}
# 1. Prepare the Data
# We filter for only the columns the model should "learn" from
ml_data <- readRDS(file.path(projPath, "data", "model_input.rds")) %>%
  # Select the target AND  features
  select(
    bike_index, med_income, med_age, pct_no_car, 
    dist_subway_miles, pop_density, pct_transit, 
    pct_walk, pct_wfh, pct_commute_60p, pct_bachelors, 
    pct_mngmt, pct_service, pct_sales, pct_const, pct_trans
  ) %>%
  drop_na()

# 2. Fit the Linear Model
# Formula: bike_index is explained by (~) all other variables (.)
lm_fit <- lm(bike_index ~ ., data = ml_data)

# 3. View the "Coefficients"
# This is the heart of the model
summary(lm_fit)

###
# Load broom for tidy model outputs
pacman::p_load(broom, dotwhisker)

# Turn the model results into a clean table
lm_results <- tidy(lm_fit) %>%
  filter(term != "(Intercept)") # Remove the intercept for better scaling

# Plot the "Pull" of each feature
ggplot(lm_results, aes(x = estimate, y = reorder(term, estimate))) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  geom_point(size = 3, color = "darkblue") +
  geom_errorbarh(aes(xmin = estimate - std.error, xmax = estimate + std.error), height = 0.2) +
  theme_minimal() +
  labs(title = "Linear Model: What drives the 'Bike Index'?",
       subtitle = "Points to the right increase biking; points to the left decrease it.",
       x = "Estimate (Coefficient)",
       y = "Feature")
```

## Misc. Notes Dump

The "Hybrid" Workflow (The Pro Move)
Since you are using GitHub, here is how you should structured your work:

data_prep.Rmd: Use this to join the Census data, TLC data, and Citi Bike data. At the end of this file, use write_rds(final_data, "data/model_input.rds").

app.R: This file should be very "slim." It just reads model_input.rds, runs the ML prediction, and shows the map.

Analogy: data_prep.Rmd is the kitchen where you chop the vegetables; app.R is the dining table where the customer sees the finished meal. You don't want the customer sitting in the middle of the kitchen! 